---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r eval = T}
library(data.table)
library(ggplot2)
library(lattice)
```

## 1. Loading and preprocessing the data
a. Load the data (i.e. read.csv())

```{r eval = T}
activity <- fread('activity/activity.csv')
```

```{r eval= T}
dim(activity)
```

```{r eval= T}
head(activity,3)
```

b. Process/transform the data (if necessary) into a format suitable for your analysis

No need at the moment

## 2. What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the data set.

a. Calculate the total number of steps taken per day
```{r eval=T}
stepsByDate <- tapply(activity$steps,activity$date,sum,na.rm=T)
stepsByDate
```

b. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r eval=T}
stepsByDate <- melt(stepsByDate)
names(stepsByDate) <- c('date','steps')
```
```{r eval=T}
hist(stepsByDate$steps,col = 'red',breaks = length(stepsByDate$steps),xlab='Steps', main ='Histogram of daily steps with NA')
```
```{r eval=T}
head(stepsByDate,3)
```

c.Calculate and report the mean and median of the total number of steps taken per day
```{r eval=T}
mean(stepsByDate$steps)
```
```{r eval=T}
median(stepsByDate$steps)
```

##### Answer: the mean of steps taken per day is 9354.23, the median is 10395

## 3. What is the average daily activity pattern?
a. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r eval=T}
stepSumByInterval <- melt(tapply(activity$steps,activity$interval,sum,na.rm=T))
names(stepSumByInterval) <- c('interval','steps')
```

get the no. days
```{r eval=T}
length(unique(activity$date))
```

However, the interval is not intuitive.

```{r eval=T}
class(stepSumByInterval$interval)
```
```{r eval=T}
stepSumByInterval$interval <- formatC(stepSumByInterval$interval, width = 4, flag='0')
head(stepSumByInterval,3)
```

Notice that the interval is in a format of hours and minutes. This is homogenous time series.
```{r eval=T}
intervalTime <- strptime(stepSumByInterval$interval,'%H%M')
intervalHour <- hour(intervalTime)
intervalMin <- minute(intervalTime)
stepSumByInterval$interval <- intervalHour*60 + intervalMin
```
```{r eval=T}
plot(stepSumByInterval$interval,stepSumByInterval$steps / 61, pch=19, type='l',col="blue",xlab = "Time in 5-minute interval", ylab="Average Number of Steps", main = "Average Steps in Intervals")
```

b. Which 5-minute interval, on average across all the days in the data set, contains the maximum number of steps?

```{r eval=T}
which.max(stepSumByInterval$steps)
```

```{r eval=T}
stepSumByInterval[104,]
```

##### Interval 515 translate to the interval starts at the 515th minute of the day, which is the interval starts at 08:35 AM

##### Answer: the maximum of average steps of 5- minute interval occured on 08:35 AM

## 4. Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

a. Calculate and report the total number of missing values in the data set (i.e. the total number of rows with NAs)

```{r eval=T}
x <- is.na(activity)
head(x)
```
```{r eval=T}
class(x)
```
```{r eval=T}
colSums(x)
```
##### Answer: There are 2304 rows of NA in the activity data set. They are missing the value of steps.

b. Devise a strategy for filling in all of the missing values in the data set. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.


Get the indices of steps which value is NA
```{r eval=T}
NAindex <- which(is.na(activity$steps))

length(NAindex)
```
```{r eval=T}
fillNA <- function(activity){
## make a copy of activity
  activityNoNA <- activity
## get the indice of missing values 
  NAindex <- which(is.na(activity$steps))
## get the sum of steps occurred by dates
  sumOfStepsByDate <- melt(tapply(activity$steps,activity$date, sum, na.rm=T))
  names(sumOfStepsByDate) <- c('dates','steps')
## only loop through the missing values, instead of all rows of activity
  for (i in 1:length(NAindex)){
    index <- NAindex[i]
    dateOftheMissingValue <- activity[index]$date
    temp <- sumOfStepsByDate[(sumOfStepsByDate$date == dateOftheMissingValue),]
    sumoftheSteps <- temp$steps
##there are 288 intervals in 1 day, the mean is the sum / no. intervals
    activityNoNA[index]$steps <- sumoftheSteps / 288
  }
  activityNoNA
}
```
c. Create a new data set that is equal to the original data set but with the missing data filled in.

```{r eval=T}
activityNoNA <- fillNA(activity)
sum(is.na(activityNoNA))
```
d. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r eval=T}
stepsByDateFilled <- melt(tapply(activityNoNA$steps, activityNoNA$date, sum))
names(stepsByDateFilled) <- c('date','steps')
summary(stepsByDateFilled)
```
```{r eval=T}
hist(stepsByDateFilled$steps,col = 'blue',breaks = length(stepsByDateFilled$steps),xlab='Steps', main ='Histogram of daily steps with no NA')
```
 
##### Answer: No difference.Because I had na.rm =T in Question 2 and it happens that all the NA exists in days that the sum of steps is zero.

Please see the following code for proof.
```{r eval=T}
x<-activity[NAindex,]
```
```{r eval=T}
unique(x$date)
```
```{r eval=T}
stepsByDate[(stepsByDate$steps == 0),]
```
Further checking a specific date to verify results.
```{r eval=T}
activity[(activity$date == '2012-10-01'),steps]
```

## 5. Are there differences in activity patterns between weekdays and weekends?

```{r eval=T}
intervalSumOfDate <- melt(tapply(activity$steps,list(activity$date,activity$interval),sum,na.rm = T))
```
```{r eval=T}
head(intervalSumOfDate,3)
```
```{r eval=T}
names(intervalSumOfDate) <- c('date','interval','steps')
activityWeekday <- weekdays(as.Date(intervalSumOfDate$date))

setWeekend <- function(activityWeekday){
  copy <-activityWeekday
  for(i in 1:length(activityWeekday)){
    if (activityWeekday[i] %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")){
      copy[i] <- 'weekday'
      }
    if (activityWeekday[i] %in% c("Saturday","Sunday")){
      copy[i] <- 'weekend'
    }
  }
  copy
}
activityWeekday <- setWeekend(activityWeekday)

intervalSumOfDate <- cbind(activityWeekday,intervalSumOfDate)
```

```{r eval=T}
head(intervalSumOfDate)
```
```{r eval=T}
stepSumByIntervalByWeekday <- melt(tapply(intervalSumOfDate$steps, list(intervalSumOfDate$activityWeekday,intervalSumOfDate$interval),sum))
names(stepSumByIntervalByWeekday) <- c("isWeekday",'interval','steps')
```
```{r eval=T}
xyplot(steps~interval|isWeekday,data=stepSumByIntervalByWeekday,type="l",scales=list(y=list(relation="free")),layout=c(1,2),xlab = 'Interval / Time of the day', ylab = 'Number of Steps', main = 'Weekend/Weekday Steps in Intervals')
```
```{r eval=T}
summary(stepSumByIntervalByWeekday[(stepSumByIntervalByWeekday$isWeekday=='weekday'),])

```
```{r eval=T}
summary(stepSumByIntervalByWeekday[(stepSumByIntervalByWeekday$isWeekday=='weekend'),])
```
##### Answer: The weekday and weekend activity patterns are different. On weekdays,subject is more active with higher mean of step. However, activity is more spead out during weekends and concentrated in the morning on weekdays.